`methyvim`: Targeted, Robust, and Model-free Differential Methylation

### 2018-08-20 paper/paper_BiocF1000.Rmd

---
title: '`methyvim`: Targeted, Robust, and Model-free Analysis of Differential
  Methylation in R'
author:
  - name: Nima S. Hejazi
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Center for Computational Biology, University of California, Berkeley
    orcid: 0000-0002-7127-2789
    email: nhejazi@berkeley.edu
  - name: Rachael V. Phillips
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Department of Statistics, University of California, Berkeley
    orcid: 0000-0003-1432-5511
abstract: |
  We present `methyvim`, an R package implementing an algorithm for the
  nonparametric estimation of the effects of exposures on DNA methylation at CpG
  sites throughout the genome, complete with straightforward statistical
  inference for such estimates. The approach leverages variable importance
  measures derived from statistical parameters arising in causal inference,
  defined in such a manner that they may be used to obtain targeted estimates of
  the relative importance of individual CpG sites with respect to a binary
  treatment assigned at the phenotype level, thereby providing a new approach to
  identifying differentially methylated positions. The procedure implemented is
  computationally efficient, incorporating a preliminary screening step to
  isolate a subset of sites for which there is cursory evidence of differential
  methylation as well as a unique multiple testing correction to control the
  False Discovery Rate with the same rigor as would be available if all sites
  were subjected to testing. This novel technique for analysis of differentially
  methylated positions provides an avenue for incorporating flexible
  state-of-the-art data-adaptive regression procedures (i.e., machine learning)
  into the estimation of differential methylation effects without the loss of
  interpretable statistical inference for the estimated quantity.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, statistical variable importance, machine learning, targeted minimum
  loss-based estimation
bibliography: references.bib
output: BiocWorkflowTools::f1000_article
---

@@

We consider an observed data structure, on a single experimental subject (e.g.,
a patient), $O = (W, A, (Y(j) : j))$, where $(Y(j) : j = 1, \ldots J)$ is the
set of CpG sites measured by the assay in question, $A \in \{0, 1\}$ represents
a binary phenotype-level treatment, and $W$ is a vector of the phenotype-level
baseline covariates that are potential confounders (e.g., age, sex). We consider
having access to measurements on a large number $J$ of CpG sites (e.g.,
$850,000$, as measured by the Illumina _MethylationEPIC_ BeadChip arrays).
Further, let $K: j \to S_j$ be a procedure that assigns to a given CpG site $j$
a set of neighbors $S_j$ -- that is, $S_j$ is a collection of indices of the
neighbors of $j$ just as $j$ indexes the full set of CpG sites; moreover, since
$Y(j)$ is the measured methylation value for a given CpG site $j$, $Y(S_j)$ is
the measured methylation values at the neighbors $S_j$ of $j$. Note that the
definition of a neighborhood $\{j, S_j\}$ is left vague so as to facilitate the
use of user-specified strategies in implementation. We consider the case of
observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where $O \sim P \in
\mathcal{M}$, which is to say that the random variable $O$ is governed by an
unknown probability distribution $P$, assumed only to reside in a nonparametric
statistical model $\mathcal{M}$ that places no restrictions on the
data-generating process.

@@

With the data structure above in hand, we let the estimand of interest be a
$j$-specific variable importance measure (VIM) $\Psi_j(P)$, which is defined
through the _true_ (and unknown) probability distribution $P$. As a motivating
example, consider the case where $\Psi_j(P)$ is
\begin{equation}\label{vim_param}
  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
\end{equation}
where $Y(S_j)$ is the subvector of $Y$ that contains the measured methylation
values of the _neighboring sites_ of $j$ (but not site $j$ itself). This
parameter of interest is a variant of the _average treatment effect_, which has
been the subject of much attention in statistical causal inference
[@holland1986statistics, @hahn1998role, @hirano2003efficient]. As a measure of
variable importance, we seek to estimate this target parameter ($\Psi_j(P)$),
which quantifies the effect of changing a treatment $A$ on the methylation
$Y(j)$ of a CpG site $j$, accounting for any potential confounding from
phenotype-level covariates $W$ and observed methylation $Y(S_j)$ at the
neighboring sites $S_j$ of $j$. We propose a procedure that estimates this
target parameter $\Psi_j(P)$ across all CpG sites of interest $j: 1, \ldots, J$.
When data-adaptive regression procedures (i.e., machine learning) are employed
in estimating $\Psi_j(P)$, this produces a nonparametric variable importance
measure of differential methylation, allowing for the identification of
differentially methylated positions (DMPs) while avoiding many assumptions
common in the use of standard parametric regression procedures. We propose
estimating $\Psi_j(P)$ via targeted minimum loss-based estimation, which allows
for data-adaptive regression procedures to be employed in a straightforward
manner using the Super Learner algorithm for ensemble modeling [@vdl2007super,
@wolpert1992stacked, @breiman1996stacked, @gruber2009targeted,
@vdl2011targeted].

@@

As a matter of practicality, we consider only estimating the target VIM
$\Psi_j(P)$ at a subset of CpG sites, so that $j: 1, \ldots J$ does not, in fact
span the full set of assayed CpG sites. In order to determine this subset of CpG
sites, we propose the use of a pre-screening procedure, which need be nothing
more than a method for differential methylation analysis that is computationally
less demanding than the method proposed here. Formally, let the pre-screening
procedure $\theta(j): j \to \{0, 1\}$, which is to say that $\theta$ takes as
input a CpG site $j$ and returns a binary decision rule of whether to include
CpG site $j$ in a subset to be evaluated further or not. As an example, one
might consider employing a classical differential methylation analysis
procedure, such as the linear modeling approach of the
[`limma`](https://bioconductor.org/packages/limma) R package
[@smyth2004linear, @robinson2014statistical], using only CpG sites that pass a
reasonable cutoff based on the linear model (e.g., magnitude of t-statistics,
p-values) for the subsequent analysis steps in the `methyvim` pipeline. Details
on the implementation of pre-screening approaches is provided in the sequel.

@@

Due to the data-adaptive nature of the regression procedures employed in
evaluating the target VIM $\Psi_j(P)$ via targeted minimum loss-based
estimation, it is possible the a given CpG site $j$ may have _too many_
neighbors $S_j$ to be controlled for in the estimation procedure. Heuristically,
the inclusion of too many neighbors when controlling for potential confounders
may lead to instability in the estimates produced. In such cases, we propose and
implement the use of a clustering technique (e.g., partitioning around medoids)
to select a _representative_ subset of neighbors. Formally, there is likely no
best choice of a specific clustering algorithm, so we leave this aspect of the
proposed algorithm as flexible for the user [@kleinberg2003impossibility]. Note
that the goal of employing a clustering procedure in `methyvim` is to obtain a
smaller but still highly representative set of neighbors $S(j)$, so as to allow
for estimates of $\Psi_j(P)$ to account for as much confounding from neighboring
sites as is allowed by the available data.

@@

Given the choice of target parameter $\Psi_j(P)$, we propose the use of targeted
minimum loss-based estimation (TMLE) to construct and evaluate an estimator
$\psi_{n, j}$ of $\Psi_j(P)$. In the case of our motivating example, where the
target VIM is based on the average treatment effect, we make use of a TML
estimator of this parameter, which has been implemented for a general case in
the [`tmle`](https://CRAN.R-project.org/package=tmle) R package
[@gruber2011tmle]; users of `methyvim` may wish to consult the documentation of
that software package as a supplement. Generally speaking, a TML estimator is
constructed from a few simple components: (1) an estimator of the propensity
score [@rosenbaum1983central], often denoted by $g(A \mid W)$; (2) an estimator
of the outcome regression, often denoted $Q(A, W)$; and (3) a targeting step
applied to the initial estimators constructed from the aforementioned components
[@gruber2009targeted]. While we describe a few of the key properties of TML
estimators in the sequel, extended technical discussion is deferred to more
comprehensive work [@vdl2011targeted, @vdl2018targeted]. In order to construct
an estimate $\psi_{n, j}$ of $\Psi_j(P)$, it is necessary to accurately estimate
to nuisance parameters, these being the propensity score ($g(A \mid W)$) and the
outcome regression ($Q(A, W)$); moreover, data-adaptive regression procedures
may be used to obtain consistent estimates of these quantities through the
creation of a stacked regression model via the Super Learner algorithm
[@wolpert1992stacked, @breiman1996stacked, @vdl2007super], which ensures
that the resulting ensemble learner satisfies important optimality properties
with respect to cross-validation [@vdl2004asymptotic, @dudoit2005asymptotics,
@vdl2003unified]. After constructing such estimates, an iterative procedure
(i.e., the targeting step) may be used to combine these individual estimates
into an estimate of the target parameter $\Psi_j(P)$.

Importantly, TML estimators are well-suited for statistical inference, having an
asymptotically normal limiting distribution [@tsiatis2007semiparametric,
@vdl2006targeted], which allows for a closed-form expression of the variance
to be derived:
\begin{equation}\label{eqn:lim_psi}
  \sqrt{n}(\psi_{n, j} - \Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2),
\end{equation}
where $\sigma_{\text{EIF}}^2$ is the variance of the efficient influence
function, with respect to a nonparametric statistical model, of the target
parameter evaluated at the observed data. Such a convenience allows for
confidence intervals and hypothesis tests to be constructed (i.e.,
$H_{0, j}: \Psi_j(P) = 0$) in a straightforward manner.

@@

Given that we seek to estimate $\Psi_j(P)$ for a possibly large number of CpG
sites $(j: 1, \ldots J)$, the need to perform corrections for multiple testing
is clear. In order to curb the potential for false discoveries, we recommend the
use of the Benjamini and Hochberg procedure for controlling the False Discovery
Rate (FDR) [@benjamini1995controlling]; however, as the proposed procedure
involves a pre-screening step, naive application of the Benjamini and Hochberg
procedure (BH) is invalid -- instead, we rely on a modification of the procedure
to control the FDR, with established theoretical guarantees when pre-screening
is employed [@tuglus2009modified]. In brief, the modified marginal Benjamini and
Hochberg procedure to control the FDR under pre-screening works by applying the
standard BH procedure to a padded vector of p-values -- that is, letting $J$ be
the number of CpG sites tested and $K$ the number of CpG sites filtered out (so
that $J + K = P$, where $P$ is the original dimension of the genomic assay), the
modified marginal BH procedure is the application of the original BH procedure
to a vector of p-values, composed of the $J$ p-values from performing $J$
hypothesis tests (of the form $H_{0, j}: \Psi_j(P) = 0$) and $K$ additional
p-values automatically set to $1$ (for the hypothesis tests not performed on
account of pre-screening). This procedure is guaranteed to control the FDR at
the same desired rate when pre-screening is performed whereas naive application
of the BH procedure fails to do so.

>>>>

---
title: '`methyvim`: Targeted, Robust, and Model-free Differential Methylation
  Analysis in R'
author:
  - name: Nima S. Hejazi
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Center for Computational Biology, University of California, Berkeley
    orcid: 0000-0002-7127-2789
    email: nhejazi@berkeley.edu
  - name: Rachael V. Phillips
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Department of Statistics, University of California, Berkeley
    orcid: 0000-0003-1432-5511
abstract: |
  We present `methyvim`, an R package implementing an algorithm for the
  nonparametric estimation of the effects of exposures on DNA methylation at CpG
  sites throughout the genome, complete with straightforward statistical
  inference for such estimates. The approach leverages variable importance
  measures derived from statistical parameters arising in causal inference,
  defined in such a manner that they may be used to obtain targeted estimates of
  the relative importance of individual CpG sites with respect to a binary
  treatment assigned at the phenotype level, thereby providing a new approach to
  identifying differentially methylated positions. The procedure implemented is
  computationally efficient, incorporating a preliminary screening step to
  isolate a subset of sites for which there is cursory evidence of differential
  methylation as well as a unique multiple testing correction to control the
  False Discovery Rate with the same rigor as would be available if all sites
  were subjected to testing. This novel technique for analysis of differentially
  methylated positions provides an avenue for incorporating flexible
  state-of-the-art data-adaptive regression procedures (i.e., machine learning)
  into the estimation of differential methylation effects without the loss of
  interpretable statistical inference for the estimated quantity.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, variable importance, machine learning, targeted loss-based
  estimation
bibliography: references.bib
output: BiocWorkflowTools::f1000_article
---

@@

We consider an observed data structure, on a single experimental subject (e.g.,
a patient), $O = (W, A, (Y(j) : j))$, where $(Y(j) : j = 1, \ldots J)$ is the
set of CpG sites measured by the assay in question, $A \in \{0, 1\}$ represents
a binary phenotype-level treatment, and $W$ is a vector of the phenotype-level
baseline covariates that are potential confounders (e.g., age, sex). We consider
having access to measurements on a large number $J$ of CpG sites (e.g.,
$850,000$, as measured by the Illumina _MethylationEPIC_ BeadChip arrays).
Further, let $K: j \to S_j$ be a procedure that assigns to a given CpG site $j$
a set of neighbors $S_j$ -- that is, $S_j$ is a collection of indices of the
neighbors of $j$ just as $j$ indexes the full set of CpG sites; moreover, since
$Y(j)$ is the measured methylation value for a given CpG site $j$, $Y(S_j)$ is
the measured methylation values at the neighbors $S_j$ of $j$. Note that the
definition of a neighborhood $\{j, S_j\}$ is left vague so as to facilitate the
use of user-specified strategies in implementation. We consider the case of
observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where $O \sim P_0
\in \mathcal{M}$, which is to say that the random variable $O$ is governed by an
unknown probability distribution $P_0$, assumed only to reside in a
nonparametric statistical model $\mathcal{M}$ that places no restrictions on the
data-generating process.

@@

With the data structure above in hand, we let the estimand of interest be a
$j$-specific variable importance measure (VIM) $\Psi_j(P_0)$, which is defined
through the _true_ (and unknown) probability distribution $P_0$. As a motivating
example, consider the case where $\Psi_j(P_0)$ is
\begin{equation}\label{vim_param}
  \Psi_j(P_0) = \mathbb{E}_{P_0}(\mathbb{E}_{P_0}(Y(j) \mid A = 1, W, Y(S_j)) -
    \mathbb{E}_{P_0} \mathbb{E}_{P_0}(Y(j) \mid A = 0, W, Y(S_j)),
\end{equation}
where $Y(S_j)$ is the subvector of $Y$ that contains the measured methylation
values of the _neighboring sites_ of $j$ (but not site $j$ itself). This
parameter of interest is a variant of the _average treatment effect_, which has
been the subject of much attention in statistical causal inference
[@holland1986statistics, @hahn1998role, @hirano2003efficient]. As a measure of
variable importance, we seek to estimate this target parameter ($\Psi_j(P_0)$),
which quantifies the effect of changing a treatment $A$ on the methylation
$Y(j)$ of a CpG site $j$, accounting for any potential confounding from
phenotype-level covariates $W$ and observed methylation $Y(S_j)$ at the
neighboring sites $S_j$ of $j$. Importantly, the use of such well-defined
parameters as variable importance measures allows the construction of
inferential procedures (e.g., confidence intervals and hypothesis tests) for the
estimate [@vdl2006statistical]. We propose a procedure that estimates this
target parameter $\Psi_j(P_0)$ across all CpG sites of interest $j: 1, \ldots,
J$. When data-adaptive regression procedures (i.e., machine learning) are
employed in estimating $\Psi_j(P_0)$, this produces a nonparametric variable
importance measure of differential methylation, allowing for the identification
of differentially methylated positions (DMPs) while avoiding many assumptions
common in the use of standard parametric regression procedures. We propose
estimating $\Psi_j(P)$ via targeted minimum loss-based estimation, which allows
for data-adaptive regression procedures to be employed in a straightforward
manner using the Super Learner algorithm for the construction of ensemble models
[@vdl2007super, @wolpert1992stacked, @breiman1996stacked, @gruber2009targeted,
@vdl2011targeted].

@@

As a matter of practicality, we consider only estimating the target VIM
$\Psi_j(P_0)$ at a subset of CpG sites, so that $j: 1, \ldots J$ does not, in
fact span the full set of assayed CpG sites. In order to determine this subset
of CpG sites, we propose the use of a pre-screening procedure, which need be
nothing more than a method for differential methylation analysis that is
computationally less demanding than the method proposed here. Formally, let the
pre-screening procedure $\theta(j): j \to \{0, 1\}$, which is to say that
$\theta$ takes as input a CpG site $j$ and returns a binary decision rule of
whether to include CpG site $j$ in a subset to be evaluated further or not. As
an example, one might consider employing a classical differential methylation
analysis procedure, such as the linear modeling approach of the
[`limma`](https://bioconductor.org/packages/limma) R package
[@smyth2004linear, @robinson2014statistical], using only CpG sites that pass a
reasonable cutoff based on the linear model (e.g., magnitude of t-statistics,
p-values) for the subsequent analysis steps in the `methyvim` pipeline.

@@

Due to the data-adaptive nature of the regression procedures employed in
evaluating the target VIM $\Psi_j(P_0)$ via targeted minimum loss-based
estimation, it is possible that a given CpG site $j$ may have _too many_
neighbors $S_j$ to be controlled for in the estimation procedure. Heuristically,
the inclusion of too many neighbors when controlling for potential confounders
may lead to instability in the estimates produced. In such cases, we propose and
implement the use of a clustering technique (e.g., partitioning around medoids)
to select a _representative_ subset of neighbors. Formally, there is likely no
best choice of a specific clustering algorithm, so we leave this aspect of the
proposed algorithm as flexible for the user [@kleinberg2003impossibility]. Note
that the goal of employing a clustering procedure in `methyvim` is to obtain a
smaller but still highly representative set of neighbors $S(j)$, so as to allow
for estimates of $\Psi_j(P_0)$ to account for as much confounding from
neighboring sites as is allowed by the available data.

@@

Given the choice of target parameter $\Psi_j(P_0)$, we propose the use of
targeted minimum loss-based estimation (TMLE) to construct and evaluate an
estimator $\psi_{n, j}$ of $\Psi_j(P_0)$. In the case of our motivating example,
where the target VIM is based on the average treatment effect, we make use of a
TML estimator of this parameter, which has been implemented for a general case
in the [`tmle`](https://CRAN.R-project.org/package=tmle) R package
[@gruber2011tmle]; users of `methyvim` may wish to consult the documentation of
that software package as a supplement. Generally speaking, a TML estimator is
constructed from a few simple components: (1) an estimator of the propensity
score [@rosenbaum1983central], often denoted $g(A \mid W)$; (2) an estimator of
the outcome regression, often denoted $\bar{Q}(A, W)$; and (3) a targeting step
that revises the initial estimators of the aforementioned components such that
the resultant estimator satisfies a set of score-like equations
[@gruber2009targeted]. While we describe a few of the key properties of TML
estimators in the sequel, extended technical discussion is deferred to more
comprehensive work [@vdl2011targeted, @vdl2018targeted]. In order to construct
an estimate $\psi_{n, j}$ of $\Psi_j(P_0)$, it is necessary to accurately
estimate two nuisance parameters, these being the propensity score
($g(A \mid W)$) and the outcome regression ($\bar{Q}(A, W)$); moreover,
data-adaptive regression procedures may be used to obtain consistent estimates
of these quantities through the creation of a stacked regression model via the
Super Learner algorithm [@wolpert1992stacked, @breiman1996stacked,
@vdl2007super], which ensures that the resulting ensemble model satisfies
important optimality properties with respect to cross-validation
[@vdl2004asymptotic, @dudoit2005asymptotics, @vdl2003unified]. To use our
running example, one would construct a TML estimator of the average treatment
effect (ATE) as a variable importance measure in the following short series of
steps:

1. Construct initial estimates of the propensity score $g(A \mid W)$ and outcome
   regression $\bar{Q}(A, W)$. As noted previously, the Super Learner algorithm
   may be employed to construct such estimates.
2. Update these initial estimates (e.g., iteratively) so as to solve the
   efficient influence function estimation equation of the parameter of interest
   relative to a nonparametric statistical model.
3. In the efficient influence function for the ATE, $\text{EIF}(P_0)(O) =
   H_n(Y - \bar{Q}(A, W))$, the auxiliary term $H_n$ may be denoted as
   $H_n = \frac{\mathbb{I}(A = 1)}{g(1 \mid W)} -
   \frac{\mathbb{I}(A = 0)}{g(0 \mid W)}$. Note that the exact form of the
   efficient influence function and auxiliary covariate varies with the target
   parameter of interest.
4. The TML estimator of the parameter of interest is constructed by a procedure
   that updates the initial estimators such that the empirical mean of the
   efficient influence function estimating equation(s) is nearly zero (with
   respect to an error proportional to the sample size, i.e.,
   $\frac{1}{\sqrt{n}}$).

Importantly, TML estimators are well-suited for statistical inference, having an
asymptotically normal limiting distribution [@tsiatis2007semiparametric,
@vdl2006targeted], which allows for a closed-form expression of the variance
to be derived:
\begin{equation}\label{eqn:lim_psi}
  \sqrt{n}(\psi_{n, j} - \Psi_j(P_0)) \sim N(0, \sigma_{\text{EIF}}^2),
\end{equation}
where $\sigma_{\text{EIF}}^2$ is the variance of the efficient influence
function, with respect to a nonparametric statistical model, of the target
parameter evaluated at the observed data. Such a convenience allows for
confidence intervals and hypothesis tests to be constructed (i.e.,
$H_{0, j}: \Psi_j(P_0) = 0$) in a straightforward manner. What's more, under
standard regularity conditions, and with consistent estimation of the propensity
score and outcome regression, the TML estimator is asymptotically efficient and
achieve the lowest possible variance among a large class of estimators
[@bickel1993efficient].

@@

Given that we seek to estimate $\Psi_j(P_0)$ for a possibly large number of CpG
sites $(j: 1, \ldots J)$, the need to perform corrections for multiple testing
is clear. In order to curb the potential for false discoveries, we recommend the
use of the Benjamini and Hochberg procedure for controlling the False Discovery
Rate (FDR) [@benjamini1995controlling]; however, as the proposed procedure
involves a pre-screening step, naive application of the Benjamini and Hochberg
procedure (BH) is invalid -- instead, we rely on a modification of the procedure
to control the FDR, with established theoretical guarantees when pre-screening
is employed [@tuglus2009modified]. In brief, the modified marginal Benjamini and
Hochberg procedure to control the FDR under pre-screening works by applying the
standard BH procedure to a padded vector of p-values -- that is, letting $J$ be
the number of CpG sites tested and $K$ the number of CpG sites filtered out (so
that $J + K = P$, where $P$ is the original dimension of the genomic assay), the
modified marginal BH procedure is the application of the original BH procedure
to a vector of p-values, composed of the $J$ p-values from performing $J$
hypothesis tests (of the form $H_{0, j}: \Psi_j(P_0) = 0$) and $K$ additional
p-values automatically set to $1$ (for the hypothesis tests not performed on
account of pre-screening). This procedure is guaranteed to control the FDR at
the same desired rate when pre-screening is performed whereas naive application
of the BH procedure fails to do so.
